# (PART) All the graph things {-} 

# Overview {-#graphics-overview}

```{r include = FALSE}
source("common.R")
```

<!--Original content: https://stat545.com/graph00_index.html-->

We work on visualization throughout the course. Here are the bits in rough order of presentation.

* Chapter \@ref(graphics-slides) - R graphics landscape slides
  - Why we prefer ggplot2 over base R graphics or lattice
  - The underappreciated importance of data frames, tidy data, and factor management to graphics
  - Basic jargon of ggplot2
* Chapter \@ref(ggplot2-tutorial) - Learning ggplot2 by using it
  - My [ggplot2 tutorial][ggplot2-tutorial] gives indicative code and all resulting figures 
  - Scatterplots, stripplots, distributions, bars, themes, managing a color scheme, bubble and line plots
* Chapter \@ref(effective-graphs) - Do's and don'ts of making effective graphs
  - Effective = easy for audience to decode numerical info
  - Our ability to decode position along common axis >> area, angle, color, etc. 
* Colors 
  - Chapter \@ref(colors) - Using colors in R
  - Chapter \@ref(qualitative-colors) Taking control of qualitative colors in ggplot2
* Practical pro tips, i.e. a return to mechanics
  - Chapter \@ref(secrets) - Secrets of happy graphing life: data.frames! tidy data! factors!
  - Chapter \@ref(save-figs) - Writing figures to file
  - Chapter \@ref(multiple-plots) - Multiple plots on a page


<!--TODO: The R Graph Catalog link and the ugrad.stat.ubc.ca link are both broken. Find a replacement? Currently links all point to github repo-->

<!--JB:
* The [R Graph Catalog](http://shiny.stat.ubc.ca/r-graph-catalog/) presents a visual, clickable index of 100+ figures
    - mostly from Naomi Robbins' book ["Creating More Effective Graphs"](http://www.amazon.com/Creating-Effective-Graphs-Naomi-Robbins/dp/0985911123)
    - see figure and the exact ggplot2 code to produce it, side-by-side
    - code for all figures and app itself is on [GitHub](https://github.com/jennybc/r-graph-catalog)
* underappreciated basics: density plots, high-volume scatterplots good stuff here: http://www.ugrad.stat.ubc.ca/~stat540/seminars/seminar03_graphics-ggplot2.html
-->



# R graphics landscape {#graphics-slides}


<script async class="speakerdeck-embed" data-id="f5ebca79660c4c3eb05b5bfbe1018545" data-ratio="1.33333333333333" src="//speakerdeck.com/assets/embed.js"></script> <div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jennybc/ggplot2-tutorial" title="hello ggplot2" target="_blank">hello ggplot2!</a> </strong> from <strong><a href="https://speakerdeck.com/jennybc" target="_blank">Jennifer Bryan</a></strong> </div>


```{r links, child="links.md"}
```