# (PART) All the package things {-} 

# Overview {-#package-overview}

```{r include = FALSE}
source("common.R")
```

<!--Original content: https://stat545.com/packages00_index.html-->

* Chapter \@ref(package-slides) - Writing your first R package slides
  + What is an R package?
  + What is a library?
  + Why make an R package?
  + How `devtools` creates a happy workflow.
* Chapter \@ref(system-prep) - Prepare your system for package development
  + Although we'll build a very simple package, we're still going to use the most modern and powerful tools for R package development. In theory, this could eventually involve compiling C/C++ code, which means you need what's called a *build environment*. See Chapter \@ref(system-prep) for help preparing your system.
* Chapter \@ref(package-from-scratch) - Write your own R Package
  + A guided walk-through of creating a small package from scratch.

## Resources {-}

* [R Packages][r-pkgs2] book: the second edition is under development by Hadley Wickham and Jennifer Bryan.
* [Writing R Extensions][cran-r-extensions], the One True Official Document on creating R packages.

```{r links, child="links.md"}
```