# (PART) Get data from the web {-} 

# Introduction {-#api-intro} 

<!--[Link to original content](https://stat545.com/webdata02_activity.html)-->

Accompanying slides are [here](http://stat545.com/webdata01_slides.html)

```{r echo = FALSE}
# Revisiting this later; issues with zoom
#knitr::include_url("http://stat545.com/webdata01_slides.html")
```


There are many ways to obtain data from the Internet; let's consider four categories:

<!--TODO: Add links to specific chapters-->

1. **click-and-download** on the internet as a "flat" file, such as .csv, .xls
2. **install-and-play** an API for which someone has written a handy R package (i.e. an API-wrapping package)
3. **API-query** published with an unwrapped API
4. **scraping** implicit in an html website


## Resources {-}

From the [rOpenSci repository of R tools for interacting with the internet](https://github.com/ropensci/webservices).

* `downloader::download()` for SSL
* `curl::curl()` for SSL.
* `httr::GET` data read this way needs to be parsed later with read.table
* `rio::import()` can "read a number of common data formats directly from an https:// URL".  Isn't that very similar to the previous?

```{r links, child="links.md"}
```
