--- 
title: "Stat545"
author: "Jenny Bryan et al."
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "This is beginning..."
---

# Stat545 {-}

## What this is

## Who is this for

## Who are we

## General advice for getting unstuck

*Also see Jenny's [slide deck on {reprex}](https://speakerdeck.com/jennybc/reprex-reproducible-examples-with-r)*

What to do when you are stuck ...

Help yourself or, __maybe__, ~~drag others into your hell~~ ask for help. The links below contain excellent self-help advice.

While you are getting started, I recommend you [seek help within the STAT545 community](http://stat545.com/index.html) first, before, e.g. posting to external forums. We are more cuddly.

Here are some good resources to get your head in the right place.

  * [Google is your friend](http://xkcd.com/627/). Google is your friend. Google is your friend.
  
    - E.g. copy the error message, maybe strip out anything highly specific, such the name of your R objects, surround with quotes and Google it!

  * This stackoverflow thread [How to make a great R reproducible example?](http://stackoverflow.com/questions/5963269/how-to-make-a-great-r-reproducible-example/5963610#5963610) is fantastic. I find that, in the process of preparing my great reproducible example, I [answer my own question](http://htmlpreview.github.io/?https://gist.githubusercontent.com/jennybc/58466fd018823302f398/raw/4c7059558e1525b482f1ee794a5828ca8999bf98/2014-07-25_wrapper-graphics-dev-open-close.html) 90% of the time. YMMV. So it is a useful exercise even if you aren't ready to call in the experts yet. The process of stripping your problem down to its bare essence often reveals where the root issue lies. __Highly recommended!__
  
  * Search [stackoverflow](http://stackoverflow.com) and include the `[r]` tag. Or the `[ggplot2]` tag. Or the `[plyr]` tag. You get the idea.

  * Check out [The R Inferno](http://www.burns-stat.com/documents/books/the-r-inferno/) (see the 9th circle: "Unhelpfully seeking help").
  
  > If someone has the wit and knowledge to answer your question, they probably have other things they would like to do. Making your message clear, concise and user-friendly gives you the best hope of at least one of those strangers diverting their attention away from their life towards your problem.
  
  * Read [How To Ask Questions The Smart Way](http://www.catb.org/~esr/faqs/smart-questions.html) by Eric Raymond and Rick Moen. You'll need a slightly thick skin, but they speak truth.
  
  > ... demonstrate the kind of attitude that leads to competence: alert, thoughtful, observant, willing to be an active partner in developing a solution
  
  * Read the [R-help posting guide](http://www.r-project.org/posting-guide.html). *But don't rush to post any questions there!*

  * More links via [How to Ask Questions](https://github.com/ireapps/first-web-scraper/blob/cb622d99305b872ebcb7bbc01fa7a97ba1cd8a4a/notes/how-to-ask-questions.md) from a web-scraping class
  
    - [How to Ask for Programming Help](http://codingkilledthecat.wordpress.com/2012/06/26/how-to-ask-for-programming-help/)
    - [Stack Exchange's 'How to ask'](https://codereview.stackexchange.com/help/how-to-ask)
    - [Matt Gemmell's 'What have you tried?'](http://mattgemmell.com/what-have-you-tried/)

## Colophon

<!--TODO: Update website and github links when finalized-->

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/). The [website](https://gracious-allen-b2d17f.netlify.com) is hosted with [netlify](https://www.netlify.com), and automatically updated after every commit by [travis-ci](https://travis-ci.org). The complete source is available from [GitHub](https://github.com/rstudio-education/stat545-reboot).


This version of the book was built with:

```{r}
library(devtools)
session_info()
```

## License {-}

<!--TODO: I was only able to get the font-awesome icons in the menu and in the text to render when I include the chunk below. There must be a better way to do this. -->
```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

This work is licensed under a [Creative Commons Attribution 4.0 International License](http://creativecommons.org/licenses/by/4.0/).

<center>
<i class="fab fa-creative-commons fa-2x"></i><i class="fab fa-creative-commons-by fa-2x"></i>
</center>

*This is a human-readable summary of (and not a substitute for) the license.
Please see <https://creativecommons.org/licenses/by/4.0/legalcode> for the full legal text.*

**You are free to:**

- **Share**---copy and redistribute the material in any medium or
  format

- **Remix**---remix, transform, and build upon the material for any
  purpose, even commercially.

The licensor cannot revoke these freedoms as long as you follow the
license terms.

**Under the following terms:**

- **Attribution**---You must give appropriate credit, provide a link
  to the license, and indicate if changes were made. You may do so in
  any reasonable manner, but not in any way that suggests the licensor
  endorses you or your use.

- **No additional restrictions**---You may not apply legal terms or
  technological measures that legally restrict others from doing
  anything the license permits.

**Notices:**

You do not have to comply with the license for elements of the
material in the public domain or where your use is permitted by an
applicable exception or limitation.

No warranties are given. The license may not give you all of the
permissions necessary for your intended use. For example, other rights
such as publicity, privacy, or moral rights may limit how you use the
material.

 Made with the <a href="https://cran.r-project.org/" target="_blank" rel="noopener"><i class="fab fa-r-project"></i></a> <a href="https://github.com/rstudio/blogdown" target="_blank" rel="noopener">blogdown</a> package and the 
    <a href="https://sourcethemes.com/academic/" target="_blank" rel="noopener">Academic theme</a> for
<a href="https://gohugo.io" target="_blank" rel="noopener">Hugo</a>.

```{r include = FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
