--- 
title: "Stat545"
author: "Jenny Bryan et al."
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Data wrangling, exploration, and analysis with R."
---

```{r include = FALSE}
source("common.R")
```

# Data wrangling, exploration, and analysis with R {-}

Learn how to:

* explore, groom, visualize, and analyze data
* make all of that reproducible, reusable, and shareable
* using R

This is a bookdown version of STAT 545 @ University of British Columbia, a course in data wrangling, exploration, and analysis with R. The contents are from the [original STAT 545 website](http://stat545.com), which is created by [this repository on GitHub](https://github.com/STAT545-UBC/STAT545-UBC.github.io). 

## Contributors {-}

Jenny Bryan ([jennybryan.org](https://jennybryan.org)), Software Engineer at [RStudio](https://www.rstudio.com/) on the [tidyverse](https://www.tidyverse.org/)/[r-lib](https://github.com/r-lib/) team. Main author.

<!--TODO: Add STAT 545 TA Alum-->

## Colophon {-}

<!--TODO: Update website and github links when finalized-->

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/). The [website](https://gracious-allen-b2d17f.netlify.com) is hosted with [netlify](https://www.netlify.com), and automatically updated after every commit by [travis-ci](https://travis-ci.org). The complete source is available from [GitHub](https://github.com/rstudio-education/stat545-reboot).


This version of the book was built with:

```{r message = FALSE}
library(devtools)
session_info()
```

## License {-}

<!--TODO: I was only able to get the font-awesome icons in the menu and in the text to render when I include the chunk below. There must be a better way to do this. -->
```{r setup, include=FALSE}
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

This work is licensed under a [Creative Commons Attribution 4.0 International License](http://creativecommons.org/licenses/by/4.0/).

<center>
<i class="fab fa-creative-commons fa-2x"></i><i class="fab fa-creative-commons-by fa-2x"></i>
</center>

*This is a human-readable summary of (and not a substitute for) the license.
Please see <https://creativecommons.org/licenses/by/4.0/legalcode> for the full legal text.*

**You are free to:**

- **Share**---copy and redistribute the material in any medium or
  format

- **Remix**---remix, transform, and build upon the material for any
  purpose, even commercially.

The licensor cannot revoke these freedoms as long as you follow the
license terms.

**Under the following terms:**

- **Attribution**---You must give appropriate credit, provide a link
  to the license, and indicate if changes were made. You may do so in
  any reasonable manner, but not in any way that suggests the licensor
  endorses you or your use.

- **No additional restrictions**---You may not apply legal terms or
  technological measures that legally restrict others from doing
  anything the license permits.

**Notices:**

You do not have to comply with the license for elements of the
material in the public domain or where your use is permitted by an
applicable exception or limitation.

No warranties are given. The license may not give you all of the
permissions necessary for your intended use. For example, other rights
such as publicity, privacy, or moral rights may limit how you use the
material.

```{r include = FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
