--- 
title: "STAT 545"
subtitle: "Data wrangling, exploration, and analysis with R"
author: 
- Jenny Bryan^[Software Engineer at [RStudio](https://www.rstudio.com/) on the [tidyverse](https://www.tidyverse.org/)/[r-lib](https://github.com/r-lib/) team, [jennybryan.org](https://jennybryan.org)]
- The STAT 545 TAs
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "STAT 545: Data wrangling, exploration, and analysis with R."
github-repo: rstudio-education/stat545
twitter-handle: jennybryan
---

# Welcome to STAT 545 {-}

```{r include = FALSE}
source("common.R")
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
library(tidyverse)
library(devtools)
library(DT)
```

Learn how to:

* explore, groom, visualize, and analyze data
* make all of that reproducible, reusable, and shareable
* using R

This is a bookdown version of STAT 545 @ University of British Columbia, a course in data wrangling, exploration, and analysis with R. 

> "The STAT 545 course became notable as an early example of a data science course taught in a statistics program. It is also notable for its focus on teaching using modern R packages, Git and GitHub, its extensive sharing of teaching materials openly online, and its strong emphasis on practical data cleaning, exploration, and visualization skills, rather than algorithms and theory." 
>  
> --- [Wikipedia](https://en.wikipedia.org/wiki/Jenny_Bryan)

The contents of this book came from the [original STAT 545 website](https://STAT545-UBC.github.io/STAT545-UBC.github.io/), which is created by [this repository on GitHub](https://github.com/STAT545-UBC/STAT545-UBC.github.io). 

## Colophon {-}

<!--TODO: Update website and github links when finalized-->

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/). The [website](https://gracious-allen-b2d17f.netlify.com) is hosted with [netlify](https://www.netlify.com), and automatically updated after every commit by [travis-ci](https://travis-ci.org). The complete source is available from [GitHub](https://github.com/rstudio-education/stat545).


This version of the book was built with:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session <- devtools::session_info()

session$platform
```

Along with these packages:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session$packages %>% 
  select(package, loadedversion, date, source) %>% 
  DT::datatable(rownames = FALSE,
                class = 'cell-border stripe',
                filter = list(position = 'top'),
                options = list(pageLength = nrow(session$packages), 
                           autoWidth = TRUE,
                           bInfo = FALSE,
                           paging = FALSE))
```

## License {-}

This work is
licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-sa/4.0/).

<center>
<i class="fab fa-creative-commons fa-2x"></i><i class="fab fa-creative-commons-by fa-2x"></i><i class="fab fa-creative-commons-sa fa-2x"></i>
</center>

*This is a human-readable summary of (and not a substitute for) the license.
Please see <https://creativecommons.org/licenses/by-sa/4.0/legalcode> for the full legal text.*

**You are free to:**

- **Share**---copy and redistribute the material in any medium or
  format

- **Remix**---remix, transform, and build upon the material for any
  purpose, even commercially.

The licensor cannot revoke these freedoms as long as you follow the
license terms.

**Under the following terms:**

- **Attribution**---You must give appropriate credit, provide a link
  to the license, and indicate if changes were made. You may do so in
  any reasonable manner, but not in any way that suggests the licensor
  endorses you or your use.
  
- **ShareAlike**---If you remix, transform, or build upon the material, you must distribute your contributions under the same license as the original. 

- **No additional restrictions**---You may not apply legal terms or
  technological measures that legally restrict others from doing
  anything the license permits.

**Notices:**

You do not have to comply with the license for elements of the
material in the public domain or where your use is permitted by an
applicable exception or limitation.

No warranties are given. The license may not give you all of the
permissions necessary for your intended use. For example, other rights
such as publicity, privacy, or moral rights may limit how you use the
material.

```{r include = FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
